---
title: "Collinearity"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Collinearity

Collinearity: where two or more predictor variables are closely related to one another. 


## Collinearity Exercises

The rank should be equal to the number of columns 

```{r }
m = matrix(c(1,1,1,1,0,0,1,1,0,1,0,1,0,0,0,1),4,4)
qr(m)$rank
```

We will visualize how there is not a single best beta-hat, when the design matrix has collinearity of columns.
```{r }
sex <- factor(rep(c("female", "male"), each=4))
trt <- factor(c("A","A","B","B","C","C","D","D"))
```

The model matrix can then be formed with:
```{r }
X <- model.matrix(~ sex + trt)
```

Check the rank
```{r }
qr(X)$rank
```
The rank is less than the number of columns

Suppose we observe some outcome, Y. For simplicity we will use synthetic data:
```{r }
Y <- 1:8
```

Now we'll construct a function which, for a given value a and b, gives us back the the sum of squared residuals after fitting the other terms.
```{r }
makeYstar <- function(a,b) Y - X[,2] * a - X[,5] * b
fitTheRest <- function(a,b) {
  Ystar <- makeYstar(a,b)
  Xrest <- X[,-c(2,5)]
  betarest <- solve(t(Xrest) %*% Xrest) %*% t(Xrest) %*% Ystar
  residuals <- Ystar - Xrest %*% betarest
  sum(residuals^2)
}
```

## Collinearity Exercises #2

What is the sum of squared residuals when the male coefficient is 1 and D coefficient is 2, and the other coefficients are fit using the linear model solution?
```{r }
fitTheRest(1,2)
```
We can apply our function fitTheRest to a grid of values for beta_male and beta_D, using the expand.grid function in R. expand.grid takes two vectors and returns a matrix with rows containing all possible combination
```{r }
expand.grid(1:3,1:3)
```
We can run fitTheRest() on a grid of values, using the following code (the Vectorize() is necessary as outer() requires only vectorized functions):
```{r }
betas = expand.grid(-2:8,-2:8)
rss = apply(betas,1,function(x) fitTheRest(x[1],x[2]))
```

## Collinearity Exercises #3
Which of the following pairs of values minimizes the RSS?
```{r }
themin= min(rss)
betas[which(rss==themin),]
```

```{r }

```