---
title: "Matrix Multiplication I"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Matrix Multiplication I

Copmuter often prefer matrices. It makes the computations faster and easier.  

## Expressing Experimental Designs

Now, we will show how to use "formula" and "model.matrix"

```{r }
x <- c(1,1,2,2)
f <- formula(~ x)
f
```

```{r }
model.matrix(f)
```

```{r }
x <- factor(c(1,1,2,2))
model.matrix(~ x)
```

```{r }
x <- factor(c(1,1,2,2,3,3))
model.matrix(~ x)
model.matrix(~ x,contrasts=list(x="contr.sum"))
```
```{r }
x <- factor(c(1,1,1,1,2,2,2,2))
y <- factor(c("a","a","b","b","a","a","b","b"))
model.matrix(~ x + y)
```
```{r }
model.matrix(~ x + y + x:y)
model.matrix(~ x*y)
```
Use function relevel to say that we want 2 to be first level

```{r}
x <- factor(c(1,1,2,2))
model.matrix(~ x)
x <- relevel(x,"2")
model.matrix(~ x)
x <- factor(x, levels=c("1","2"))
```

```{r, echo = FALSE}
z <- 1:4 
model.matrix(~ z)
model.matrix(~ 0 + z)
model.matrix(~ z + I(z^2))
```

## Linear Models in Practice I

Comparison between two groups of mice that were given two diets

```{r }
url <- "https://raw.githubusercontent.com/genomicsclass/dagdata/master/inst/extdata/femaleMiceWeights.csv"
filename <- "femaleMiceWeights.csv"
library(downloader)
if (!file.exists(filename)) download(url,filename)
set.seed(1)
dat <- read.csv("femaleMiceWeights.csv") ##previously downloaded
stripchart(dat$Bodyweight ~ dat$Diet, vertical=TRUE, method="jitter",
           main="Bodyweight over Diet")
```
## A linear model with one variable 

Analyzing the difference between the two groups using linear models

-Diet is a factor that has two levels, chow and hf. factor used to represent categorical variables in R. 
-level : used to display the unique levels of factor. 
-Then we will create matrix X 
if diet = hf we will get 1 
```{r }
dat$Diet <- as.factor(dat$Diet)
levels(dat$Diet) 
X <- model.matrix(~Diet, data = dat)
X 
```
```{r }
colnames(X)
dat$Diet <- relevel(dat$Diet, ref = "hf")
model.matrix(~Diet, data = dat)
dat$Diet <- relevel(dat$Diet, ref = "chow")
```

## Linear Models in Practice II

## Running the linear model

lm : function for linear model

```{r }
fit <- lm(Bodyweight ~ Diet, data=dat)
summary(fit)
(coefs <- coef(fit))
```
## The mathmatics behind lm()

$$ \hat{\boldsymbol{\beta}} = (\mathbf{X}^\top \mathbf{X})^{-1} \mathbf{X}^\top \mathbf{Y} $$

```{r }
Y <- dat$Bodyweight
X <- model.matrix(~ Diet, data=dat)
solve(t(X) %*% X) %*% t(X) %*% Y
```

Another way using the average: 
```{r }
s <- split(dat$Bodyweight, dat$Diet)
mean(s[["chow"]])
mean(s[["hf"]]) - mean(s[["chow"]])
```