---
title: "Matrix Algebra"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

In this notebook, we will show how to create vectors and matrices. Also, we will demonstrate matrix operations.


scaler : numbers 
vector: series of numbers. we use in R: c(1,5,3,4)
matrix: series of vectors
we can turn vectors into matrices using: rbind() , cbind() , matrix()

```{r cars}
vector = c(1,5,3,4)
X = matrix(1:1000, 100,10)
```

## Matrix Notation Exercises #2

```{r , echo=FALSE}
first_col <- 1:10
y <- cbind(x1=first_col, x2=first_col*2,x3=first_col*3,x4=first_col*4,x5=first_col*5)
sum(y[7,])
```

## Matrix Operations

use t() for the transpose 

```{r }
X <- matrix(1:12,4,3)
print(X)
print(X *2)
t(X)

```



```{r}
X <- matrix(c(1,3,2,1,-2,1,1,1,-1),3,3)
y <- matrix(c(6,2,1),3,1)
solve(X) %*%y # or use solve(x,y)

```
# Matrix Operation Exercises #2
```{r}
X <- matrix(c(3,2,1,5,4,2,-1,0,-5,2,5,0,1,-1,-5,1),4,4)
y <- matrix(c(10,5,7,4),4,1)
solve(X,y)

```

# Matrix Operation Exercises #3

```{r}
a <- matrix(1:12, nrow=4)
b <- matrix(1:15, nrow=3)
# What is the value in the 3rd row and the 2nd column of the matrix product of a and b?
c <- a %*% b # multiplication
c[3,2]

```

# Matrix Operation Exercises #4
Multiply the 3rd row of a with the 2nd column of b, using the element-wise vector multiplication with *.

```{r}
sum(a[3,] * b[,2])
```

# The sample mean
We can obtain Ybar by using two methods 

```{r }
library(UsingR)
y <- father.son$sheight
print(mean(y))
N <- length(y)
Y <- matrix(y,N,1)
A <- matrix(1,N,1)
barY <- t(A) %*%Y / N #first way
barY <- crossprod(A,Y) / N #second way
print(barY)
```

## The Variance

```{r}
r <- y - barY
crossprod(r) / N
# below is the same
var(y) * (N-1) / N
```
