---
title: "QR"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## QR

Use QR factorization when we can't solve X to minimize rss
X = QR
Q a N x p matrix with Q^t * Q = I
R a P x p upper triangular matrix

## Example
```{r }
n <- 50;M <- 500
x <- seq(1,M,len=n)
X <- cbind(1,x,x^2,x^3)
beta <- matrix(c(1,1,1,1),4,1)
y <- X%*%beta+rnorm(n,sd=1)
solve(crossprod(X))
```
we see that we can't solve crossprod(X)
Now we will use QR
```{r }
QR <- qr(X)
Q <- qr.Q(QR)
R <- qr.R(QR)
betahat <- backsolve(R, crossprod(Q,y))
betahat
```
There is a function in R can do all that 
```{r }
QR <- qr(X)
betahat <- solve.qr(QR, y)
```

```{r }
fitted <- tcrossprod(Q)%*%y
lines(x,fitted,col=2)
```
We can compute Standard Errors
```{r }
df <- length(y) - QR$rank
sigma2 <- sum((y-fitted)^2)/df
var <- sigma2*chol2inv(QR$qr)
SE <- sqrt(diag(var))
cbind(betahat, SE)
```
Use linear model
```{r }
summary(lm(y~X-1))$coef
```
## QR Exercises
```{r }
fit <- lm(friction ~ type + leg, data=spider)
betahat <- coef(fit)
Y <- matrix(spider$friction, ncol=1)
X <- model.matrix(~ type + leg, data=spider)
```

## QR Exercises #1
What is the first row, first column element in the Q matrix for this linear model?
```{r }
QR <- qr(X)
Q <- qr.Q(QR)
t(Q[1,1])
```

What is the first row, first column element in the R matrix for this linear model?
```{r }
R <- qr.R( QR )
t(R[1,1])
```

What is the first row, first column element of Q^t * Y?
```{r }
Rbetahat <- crossprod(Q,Y)
Rbetahat[1,1]
```
