---
title: "Minimizing the RSS"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

Residual Sum of Squares (RSS) is the sum of the squared differences between the observed values (actual responses) and the values predicted by the regression model. 

linear model example : y = beta0 + beta1 (X1) + c 

Ordinary Least Squares (OLS) method used to estimate the coefficients in linear regression (beta). 
The OLS solution provides the coefficients that minimize the RSS, resulting in a regression line that best fits the observed data.

```{r }
library(UsingR)
x=father.son$fheight
y=father.son$sheight
X <- cbind(1,x)
betahat <- solve(t(X)%*%X)%*%t(X)%*%y
betahat <- solve(crossprod(X))%*%crossprod(X,y) 
```

## Plot

```{r pressure, echo=FALSE}
g <- 9.8 
n <- 25
tt <- seq(0,3.4,len=n) 
f <- 56.67  - 0.5*g*tt^2
y <-  f + rnorm(n,sd=1)
plot(tt,y,ylab="Distance in meters",xlab="Time in seconds")
lines(tt,f,col=2)
```

## Plot the minimum

```{r pressure, echo=FALSE}
rss <- function(Beta0,Beta1,Beta2){
  r <- y - (Beta0+Beta1*tt+Beta2*tt^2)
  return(sum(r^2))
}
Beta2s <- seq(-10,0,len=100)
RSS <- sapply(Beta2s, rss, Beta0=55, Beta1=0)
plot(Beta2s, RSS, type="l")
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.


## Matrix Algebra Examples Exercises
X <- matrix(c(1,1,1,1,0,0,1,1),nrow=4)
rownames(X) <- c("a","a","b","b")
Suppose that the fitted parameters for a linear model
beta <- c(5, 2)

# Question 1

What is the fitted value for the A samples? (The fitted Y values.)

```{r }
beta <- matrix(c(5,2),nrow=2,ncol=1)
X[1:2,] %*% beta
```

#Question 2

What is the fitted value for the B samples? (The fitted Y values.)

```{r }
X[3:4,] %*% beta
```


## Inference Review 

#Exercises #1

```{r }
RNGkind("Mersenne-Twister", "Inversion", "Rejection")
g = 9.8 
h0 = 56.67
v0 = 0
n = 25
tt = seq(0,3.4,len=n) ##time in secs, t is a base function
y = h0 + v0 *tt  - 0.5* g*tt^2 + rnorm(n,sd=1)

X = cbind(1,tt,tt^2)
A = solve(crossprod(X))%*%t(X)
```


# Exercises #2
```{r }
set.seed(1)
B = 100000
g = 9.8
n =25
tt = seq(0,3.4,len=n)
X = cbind(1,tt,tt^2) 
A = solve(crossprod(X))%*%t(X)
betahat = replicate(B,{
  y=56.67 - 0.5*g*tt^2 + rnorm(n,sd=1)
  betahats = -2*A%*%y
  return(betahats[3])
  })
sqrt(mean((betahat-mean(betahat))^2))
```

# Standard Errors Exercises

```{r }
library(UsingR)
x = father.son$fheight
y = father.son$sheight
n = length(y)
N = 50
set.seed(1)
index = sample(n,N)
sampledat = father.son[index,]
x = sampledat$fheight
y = sampledat$sheight
betahat = lm(y~x)$coef
```


#Standard Errors Exercises #1

fitted values Yhat from a linear model can be obtained with:
fit = lm(y~x)
fit$fitted.values

What is the sum of the squared residuals (where residuals are given by ri = yi - yhat

```{r }
sum((y - fit$fitted.values)^2)
```

#Standard Errors Exercises #2
X = cbind(rep(1,N),x)

The inverse of X transpose times X. Use the solve() function for the inverse and t() for the transpose. What is the element in the first row, first column?

```{r }
a = solve(t(X)%*% X)
a[1,1]
```

#Standard Errors Exercises #3

```{r }
fit = lm(y ~x)
sigma2 = sum((y - fit$fitted.values)^2) / (N-2)
sqrt(sigma2 * diag(solve(t(X) %*% X)))
```

# Week 2 Quiz

RNGkind("Mersenne-Twister", "Inversion", "Rejection")
X <- matrix(c(1,1,1,1,1,1,0,0,1,1,0,0,0,0,0,0,1,1),nrow=6)
rownames(X) <- c("a","a","b","b","c","c")
beta <- c(10,3,-3)

#Question #1

What is the fitted value for the B samples?
```{r }
X[3:4,] %*% beta
```

#Question #2

What is the fitted value for the C samples?
```{r }
X[5:6,] %*% beta
```

#Question #3
```{r}
library(UsingR)
x = father.son$fheight
y = father.son$sheight
n = length(y)
```

Now let's run a Monte Carlo simulation in which we take a sample of size 50 over and over again. Here is how we obtain one sample:
```{r}
N =  50
index = sample(n,N)
sampledat = father.son[index,]
x = sampledat$fheight
y = sampledat$sheight
betahat =  lm(y~x)$coef
```

Use the function replicate() to take 10,000 samples
```{r}
N = 50
B = 10000
set.seed(1)
betahat = replicate(B,{
  index = sample(n,N)
  sampledat = father.son[index,]
  x = sampledat$fheight
  y = sampledat$sheight
  lm(y~x)$coef[2]
})
sqrt ( mean( (betahat - mean(betahat) )^2 ))
```

# Question #4
```{r}
mean((y - mean(y))*(x-mean(x)))
```